---
title: "Life Expectancy"

output: 
  flexdashboard::flex_dashboard:
    theme: sandstone
---


```{r setup, include=FALSE}
library(flexdashboard)
library(gapminder)
library(ggplot2)
library(plotly)

library(sqldf)

```

 

```{r}
df <- gapminder
df <- as.data.frame(df)
```





1952 Life Expectancy Disparity
==================
Comments :

- All 5 countries with a good life expectancy are developed European countries

- All 5 countries with a very low life expectancy are third world countries, four of them being from Africa

- The highest life expectancy is 73 while the lowest is 29, that is a 44 years difference


```{r}
life_exp_1952 = df[df$year ==1952,]

```

Column {data-width=650}
-----------------------------------------------------------------------
```{r}
sorted_life_exp_1952 = life_exp_1952[order(life_exp_1952$lifeExp),]

plot_sorted_life_exp_1952 <- sorted_life_exp_1952[c(1:5, 138:142) ,]

plot_sorted_life_exp_1952

```

```{r}

# customizing  plotly tooltip 
Country = reorder(plot_sorted_life_exp_1952$country,plot_sorted_life_exp_1952$lifeExp) 
Life_Expectancy = plot_sorted_life_exp_1952$lifeExp 

# ggplot

plot_1 <- ggplot(data =plot_sorted_life_exp_1952, 
                 )+
  geom_bar(aes(x = Country,y=Life_Expectancy, fill=Life_Expectancy), stat='identity',)+
  scale_fill_continuous(low="olivedrab2", high="olivedrab4")+
  xlab('Country')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) #rotate x axis

#converting to an interactive plot
plotly_1 = ggplotly(plot_1, tooltip = c('x','y')) 


```

Column {data-width=650}
-----------------------------------------------------------------------

### 1952 Life Expectancy Disparity (interactive chart)

```{r}
plotly_1
```


2007 Life Expectancy Disparity
==================
Comment :

- With the advancement in Science and Technology the life expectancy increased overall compared to 1952

- Unlike the 50's, the top 5 countries with the highest life expectancy are not all European. 2 of them are from are from Asia, 2 from Europe, and Australia 

- Although life expectancy has doubled in Africa, all 5 countries with the lowest life expectancy are from Africa

```{r}
life_exp_2007 = df[df$year ==2007,]

```

Column {data-width=650}
-----------------------------------------------------------------------
```{r}
sorted_life_exp_2007 = life_exp_2007[order(life_exp_2007$lifeExp),]

plot_sorted_life_exp_2007 <- sorted_life_exp_2007[c(1:5, 138:142) ,]

plot_sorted_life_exp_2007

```
```{r}

# customizing  plotly tooltip 
Country = reorder(plot_sorted_life_exp_2007$country,plot_sorted_life_exp_2007$lifeExp) 
Life_Expectancy = plot_sorted_life_exp_2007$lifeExp 

# ggplot

plot_2 <- ggplot(data =plot_sorted_life_exp_2007, 
                 )+
  geom_bar(aes(x = Country,y=Life_Expectancy, fill=Life_Expectancy), stat='identity',)+
  scale_fill_continuous(low="olivedrab2", high="olivedrab4")+
  xlab('Country')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) #rotate x axis

#converting to an interactive plot
plotly_2 = ggplotly(plot_2, tooltip = c('x','y')) 


```

Column {data-width=650}
-----------------------------------------------------------------------

### 2007 Life Expectancy Disparity (interactive chart)

```{r}
plotly_2
```



Top 5 Countries (population)
==================

Comments :

- All countries life expectancy rose steadily from 1952 to 2007 

- China life expectancy decreased significantly from 1958 up until 1962. This sharp drop is due to The Great Chinese Famine that lasted 3 years(1959 and 1961 ) (source: https://en.wikipedia.org/wiki/Great_Chinese_Famine)

```{r}

df_2007_filter = df[df$year==2007,]
```






Column {data-width=450}
-----------------------------------------------------------------------
```{r}
top_5 <- df[df$country
   %in%
     list(
          df_2007_filter[
                order(-df_2007_filter$pop),][1:5,][,'country'])[[1]],]

top_5
```


Column {data-width=850}
-----------------------------------------------------------------------

### Life expectancy change (interactive chart)

```{r}
plot_3 <- ggplot(data=top_5 )+
  geom_line(aes(x=year,y=lifeExp, group=country, color=country))+
  geom_point(aes(x=year,y=lifeExp, group=country,), size=0.09)+
  scale_color_manual(values=c('Orange3','Red3','Green3','Yellow3','Blue3')) 

ggplotly(plot_3)
```

Continents life expectancy change  (1952-2007)
==================

Comments :


As you can see from the plot below, life expectant has steadily increase for all 5 continents. 
It's also apparent the richer the continent the higher the life expectancy. 

```{r}

df1 = df[df$year==2007,]

pop_2007 <- df1[order(-df1$pop),][1:5]
```





```{r}
continents <- sqldf("SELECT continent,year,pop, AVG(lifeExp) as lifeExp 
      FROM df
      GROUP BY year,continent
    ")

```

```{r}

Continent = as.factor(continents$continent)
plot_4 <- ggplot(data = continents)+
  geom_line(aes(x=year, y= lifeExp , group = Continent,color=Continent))+
  
  geom_point(aes(x=year, y= lifeExp,group = Continent), size=0.09)+
  scale_color_manual(values=c('Green3','Yellow3','Blue3','Red3','Orange3')) 

plotly_4 = ggplotly(plot_4, tooltip = c('x','y','group'))

  
```



Column {data-width=450}
-----------------------------------------------------------------------
```{r}
continents
```


Column {data-width=850}
-----------------------------------------------------------------------

### life expectancy change  (1952-2007) (interactive chart)

```{r}
plotly_4
```





