---
title: 'Project #2 (Visualizing Kiggle 2020 Survey Data)'
author: "Ivan Bayingana"
date: "1/31/2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r,warning=FALSE,message=FALSE}
library(scales)
library(data.table)
library(ggplot2)
library(sqldf)
```
```{r}
df <- read.csv("kaggle_survey_2020_responses.csv",na.strings=c("","NA"))
```
# Cleaning 


```{r}
# the first row contains the questions, let's assign it to its own variable since...
#...it's not part of the responses 
questions = df[1,]
df = df[-1,]
questions[1,2]
```

# Visuals


## 1. Age distrubution 
```{r}
ggplot(df, aes(Q1)) + 
  geom_bar(stat = "count",fill='wheat4')+
  labs(x='Age',y='Count',title = 'Age Distribution',subtitle=questions[1,2]) +
  theme(plot.title=element_text(hjust=0.5),plot.subtitle=element_text(hjust=0.5,size=9))
  

```

It appears most people who filled out this survey are between the age of 22 and 29. The older people are the less likely they are of filling out this survey.

## 2. Gender 

```{r}
ggplot(df[df$Q2=='Man' | df$Q2=='Woman',], aes(Q2)) +
geom_bar(fill='wheat4')+
labs(x='Gender',y='Count',title = 'Gender Distribution',subtitle=questions[1,3]) +
theme(plot.title=element_text(hjust=0.5),plot.subtitle=element_text(hjust=0.5,size=9))+
  geom_text(stat = "count",aes(label=after_stat(count)), vjust = -0.5) 
```


Looks like more men filled out this survey. Women account for less than 5000 survey respondents compare that to 15000 men, you can see that there is a big difference.


## 3. Country

Plotting all the countries would result in an overcrowded graph, I'm only going to plot the top 10 countries instead. 

```{r}
df_country = sqldf("select Q3 as Country, count(Q3) as Count
                   from df 
                   where Q3 != 'Other' 
                   group by Q3 
                   order by Count DESC
                   limit 10 ")


# I abbreviated the names because they were too long.
df_country[2,1] = "USA"
df_country[6,1] = "UK"
df_country
```

```{r}
ggplot(df_country, aes(x=Country, y=Count)) + 
geom_bar(stat = "summary",fun=sum)+
stat_summary( fun=sum,geom="bar",fill='wheat4')+
labs(x='Country',y='Count',title = 'Top 10 Countries',subtitle=questions[1,4])+
  theme(plot.title=element_text(hjust=0.5),plot.subtitle=element_text(hjust=0.5,size=9))+
geom_text(aes(label = Count), vjust = -0.5,color='wheat4')

  
```

India has by far the most survey respondents followed by the USA.  All other countries have less than 700 survey respondents .

## 4. Education 
```{r}
library(stringr)

```

```{r warning=FALSE,message=FALSE}
ggplot(df, aes(x=reorder(Q4,Q4,function(x)-length(x)))) +
geom_bar(stat = "count",fill='wheat4')+
labs(x='Degree',y='Count',title = 'Education',subtitle=questions[1,5]) +
theme(plot.title=element_text(hjust=0.5),
      plot.subtitle=element_text(hjust=0.5,size=9),
      axis.text.x = element_text(face = "bold",angle =45,hjust = 1))+
scale_x_discrete(na.translate = FALSE)+
coord_flip()

```

Most respondents have at least a Bachelor or Masters degree, very few have no formal education past high school. 


## 5. Employment

```{r warning=FALSE,message=FALSE}
ggplot(df, aes(Q5)) + geom_bar(stat = "count",fill='wheat4')+
labs(x='Position',y='Count',title = 'Employment',subtitle=questions[1,6]) +
  theme(plot.title=element_text(hjust=0.5),
        plot.subtitle=element_text(hjust=0.5,size=9))+
scale_x_discrete(na.translate = FALSE)+
coord_flip()
```

We see that most respondents are students which make sense considering the age distribution graph. 


## 6.Experience
```{r warning=FALSE,message=FALSE }
ggplot(df, aes(x=reorder(Q6,Q6,function(x)-length(x)))) + 
geom_bar(stat = "count",fill='wheat4')+
labs(x='Years',y='Count',title = 'Experience',subtitle=questions[1,7])+
theme(plot.title=element_text(hjust=0.5),plot.subtitle=element_text(hjust=0.5,size=9))+
scale_x_discrete(na.translate = FALSE)+
coord_flip()
```

Here the graph shows most people have less than 5 years of experience, again this makes sense considering the age distribution graph shows the majority of respondents are between the age of 22 and 29. 

## 7. programming languages


```{r}

Program_lang = sqldf("select count(Q7_Part_1) as Python,
                     count(Q7_Part_2) as R,
                     count(Q7_Part_3) as SQL,
                     count(Q7_Part_4) as C,
                     count(Q7_Part_5) as 'C++',
                     count(Q7_Part_6) as java,
                     count(Q7_Part_7) as Javascript,
                     count(Q7_Part_8) as Julia,
                     count(Q7_Part_9) as Swift,
                     count(Q7_Part_10) as Bash,
                     count(Q7_Part_11) as Matlab,
                     count(Q7_Part_12) as None,
                     count(Q7_OTHER) as Other from df")

programming_languages = transpose(Program_lang,keep.names="Language")

programming_languages

```


```{r}
ggplot(programming_languages, aes(x=Language, y=V1)) + 
  geom_bar(stat = "summary",fun=sum)+
  stat_summary( fun=sum,geom="bar",fill='wheat4')+
  
labs(x='Language',y='Count',title = 'Programming Language',subtitle='What programming 
     languages do you use on a regular basis?')+
theme(plot.title=element_text(hjust=0.5),
        plot.subtitle=element_text(hjust=0.5,size=9))+
coord_flip()

```

Python is by far the most used programming language compared to everything else available, SQL comes in second place followed by R. It will be interesting to see if some of the new programming language like Julia can catch up to the popularity of Python. 

## 8. Recommendation

```{r warning=FALSE,message=FALSE}
ggplot(df, aes(Q8)) + geom_bar(stat = "count",fill='wheat4')+
  labs(x='Language',y='Count',title = 'Recommendation',subtitle=questions[1,21])+
  theme(plot.title=element_text(hjust=0.5),plot.subtitle=element_text(hjust=0.5,size=9))+
  scale_x_discrete(na.translate = FALSE)+
  coord_flip()

```

The popularity of python is undeniable here, almost all people in this survey recommend aspiring data scientist to learn  python first. 


## 9. Environment 


```{r}
query = sqldf("select count(Q9_Part_1) as Jupyter, 
              count(Q9_Part_2) as RStudio ,
              count(Q9_Part_3) as 'Visual Studio',
              count(Q9_Part_4) as VSCode,
              count(Q9_Part_5) as 'PyCharm',
              count(Q9_Part_6) as Spyder,
              count(Q9_Part_7) as 'Notepad++'  ,
              count(Q9_Part_8) as 'Sublime Text'  ,
              count(Q9_Part_9) as  'Vim / Emacs' ,
              count(Q9_Part_10) as 'MATLAB' ,
              count(Q9_Part_11) as None,
              count(Q9_OTHER) as Other from df")

env = transpose(query,keep.names ='Environment')
env
```

```{r}
ggplot(env, aes(x=Environment, y=V1)) + 
  geom_bar(stat = "summary",fun=sum)+
  stat_summary( fun=sum,geom="bar",fill='wheat4')+
  
  labs(x='Language',y='Count',title = 'IDE',subtitle="Which of the following integrated
       development environments (IDE's) do you use on a regular basis?")+ 
  theme(plot.title=element_text(hjust=0.5),plot.subtitle=element_text(hjust=0.5,size=9))+
  coord_flip()
```

It looks like most people use Jupyter as their IDE on a regular basis. 


## 10. Notebooks

```{r}

query = sqldf("select count(Q10_Part_1) as  Kaggle , 
              count(Q10_Part_2) as 'Colab Notebooks' ,
              count(Q10_Part_3) as 'Azure Notebooks',
              count(Q10_Part_4) as  'Paperspace'
              ,count(Q10_Part_5) as ' Binder',
              count(Q10_Part_6) as 'Code Ocean',
              count(Q10_Part_7) as ' IBM Studio'  
              ,count(Q10_Part_8) as ' Amazon Studio'  ,
              count(Q10_Part_9) as  'Amazon Notebooks' ,
              count(Q10_Part_10) as 'Google AI Platform' ,
              count(Q10_Part_11) as 'Google Datalab',
              count(Q10_Part_12) as ' Databricks colab',
              count(Q10_Part_13) as None,
              count(Q10_OTHER) as Other from df")

notebooks = transpose(query,keep.names ='Notebooks')
notebooks
```
```{r}
ggplot(notebooks, aes(x=Notebooks, y=V1)) + 
  geom_bar(stat = "summary",fun=sum)+
  stat_summary( fun=sum,geom="bar",fill='wheat4')+
  
  labs(x='Notebooks',y='Count',title = 'Notebooks',subtitle="Which of the following hosted 
       notebook products do you use on a regular basis?")+
  theme(plot.title=element_text(hjust=0.5),plot.subtitle=element_text(hjust=0.5,size=9))+ 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))+
  coord_flip()
```

Unsurprisingly, Colab Notebooks and Kaggle which use a jupyter IDE are the most used hosted notebook products. 



#                                                        END

```{r warning=FALSE,message=FALSE,results='hide',echo=FALSE,fig.show='hide'}
n=2
for (i in names(df[2:20])) {
      print( ggplot(df, aes(x = reorder(unlist(df[i]),unlist(df[i]),function(x)-length(x)))) + geom_bar(fill='wheat4')+
      labs(x='x',y='Count',title=questions[1,n])+
      #scale_y_continuous(limits=c(0,125000),oob=rescale_none)+
      geom_text(stat = "count",aes(label=after_stat(count)), vjust = -0.5,color='wheat4')  + 
      scale_x_discrete(guide = guide_axis(n.dodge=3))+
      scale_x_discrete(na.translate = FALSE)+
      theme(plot.title=element_text(hjust=0.5,size=9)))
 n=n+1
 }
```

